\section{Build environment} \label{sec:080dev:build_env}

The documentation for writing applications using LibTooling mainly concerns writing tools as part of the LLVM project repository \cite{clangMatchingClangAST, clangClangTransformerTutorial}. While this is good for contributing to the project, it is not ideal for version control and developing stand-alone projects.
It was necessary to create a build environment that allowed for out-of-tree builds which utilize LibTooling. A similar attempt has been made before but the project was abandoned in 2020 \cite{kasmisClangOutoftreeBuild2023}. LLVM has since moved from a distributed repository architecture to a monolithic one marking most of the earlier attempt obsolete.
The following section is dedicated to describing the important decisions made related to the build environment.

\subsection{Build settings}

Initially, some general settings for the project are configured which can be seen in \cref{code:080dev:cmake_base_settings}. LibTooling was compiled with the Clang compiler.
Choosing another compiler may result in parts of the project being compiled with another standard library implementation, e.g., libstd++ (the default for GCC). This may cause incompatibility between the application binary interfaces (ABIs) which is considered undefined behaviour, essentially leaving the entire program behaviour unspecified \cite{cppreferenceUndefinedBehaviorCppreference}. This concept is also known as ABI breakage.
As such line 1 forces the project to be compiled with the Clang compiler.
Line 2 defines the C++ standard version, which is set to C++17 since LibTooling was compiled with this.
Line 3 defines the output directory of the executable to be in \textinline{<build_folder>/bin} which has importance concerning how LibTooling searches for include directories at run-time as described in \cref{subsec:080dev:rt_include}.
Finally, line 4 disables Run-Time Type Information (RTTI). RTTI allows the program to identify the type of an object at runtime by enabling methods such as \cppinline{dynamic_cast} and \cppinline{typeid} among others. When compiling LLVM it is up to the user whether RTTI should be included or not. It is disabled by default as it slows down the resulting executable considerably. This setting and the resulting runtime penalties are propagated to the specified subprojects such as Clang and LibTooling. In contrast, the default in CMake projects is to enable RTTI. Additionally, CMake assumes the specified libraries were also compiled with RTTI. This will result in nasty linker errors and RTTI is therefore explicitly disabled in the tool project.

\begin{listing}[H]
    \begin{cmakecode}
set(CMAKE_CXX_COMPILER clang++)
set(CMAKE_CXX_STANDARD 17)
set(CMAKE_RUNTIME_OUTPUT_DIRECTORY "${CMAKE_BINARY_DIR}/bin")
add_compile_options(-fno-rtti)
    \end{cmakecode}
    \caption{General settings for the CMake build environment.}
    \label{code:080dev:cmake_base_settings}
\end{listing}

\subsection{Run-time include directories}\label{subsec:080dev:rt_include}
When executing binaries created with LibTooling, a big part of the process is the analysis of the target source-code. The analysis is done following the pipeline as shown in \cref{fig:040des:llvmToolchainOverview}. Most projects written in C++ make use of the C++ standard library which implements many commonly used functionalities. Naturally, the tool needs to know the definitions in the standard library in order to analyse the target source-code. 
For practical reasons, LibTooling provides a mechanism for the automatic discovery of the standard library header files which should be included when parsing source files. The headers are located by using a relative path with the pattern \textinline{../lib/clang/<std_version>/include} from the location of the binary. Where \textinline{<std_version>} indicates which version of the standard library which the tool was compiled with (in this project it was 17).

This hard-coded approach is quite simple but limited, as it forces the users to only run the tool in a directory where the headers can be found in the relative directory \textinline{<current_dir>/../lib/clang/17/include}.
If the user attempts to run it somewhere else, and the analyzed files make use of standard library features, they will get an error while parsing the files (e.g. that the header \textinline{<stddef.h>} was not found).
This issue makes it more difficult to write truly independent tools as they still need some reference to the Clang headers, which would essentially mean moving the executable to the directory where Clang was compiled.

One existing solution is to provide the location of the headers as an argument to the binary when executed. This is possible since tools written with LibTooling invoke the parser of Clang, from where it is possible to forward the include directory as an argument to the compiler e.g. by specifying\\\textinline{-- -I"/usr/local/lib/clang/17"}.
However, this was found to be impractical since the location of the include path may vary depending on the system and forgetting to write the path results in errors that can be very difficult to decipher.
Instead, it was decided to create a build environment where the user must provide the location of the Clang headers when configuring CMake. Failure to do so will result in the generation of an appropriate error message. Through CMake, the necessary headers are then copied to the build directory.

The solution is by no means perfect, as the user is still forced to execute the binary from the build directory. In many situations, this is sufficient, as most IDEs follow this behaviour by default and it allows the projects to be built out-of-tree. If the user wishes to run the binary from outside the build directory, they still have the option of specifying the location through the \textinline{-- -I"<clang_include>"} option. The solution can be found in the \href{https://github.com/mortenhaahr/RD/blob/main/examples/c_style_array_converter/cmake/functions.cmake}{functions.cmake} file.

In the future, it may be desirable to explore a solution using the LLVM command line library to search some commonly used directories for the Clang headers.

\subsection{Configuring the target project}

The target project, whose source files will be analysed by the tool, should generate a compilation database. A compilation database contains information about the compilation commands invoked to build each source file. This file is used by LibTooling to detect the compile commands and include files necessary to generate the correct AST \cite{clangJSONCompilationDatabase}. Tools like CMake can auto-generate the compilation commands at configuration time. When using CMake the compilation database generation can be enabled by inserting \cmakeinline{set(CMAKE_EXPORT_COMPILE_COMMANDS 1)} in the CMakeLists.txt file.