% Discuss the difficulties of finding the correct node types - this requires mostly knowledge of how the AST tree is build, aka which node types exist and what data they contain.

\section{Finding the correct node types and data methods}

Much of the development time for the CStyle array conversion tool and the Enum to\_string generation tool was spent looking through the different AST node types defined in the LibTooling framework and determining which nodes contain what data. This was a long process during the development phase of the project, as there are many different class specializations and almost identical node types with vastly different data. One example of these difficulties was the extraction of the qualifier for the \cppinline{DeclaratorDecl} in the CStyle array conversion tool. The qualifier is the explicit namespace in front of the variable declaration e.g., the qualifier of \cppinline{A::B::C} would be \cppinline{A::B::}. This was eventually extracted directly from the source code through the  \cppinline{getQualifierLoc} method implemented for \cppinline{DeclaratorDecl} nodes, but many other options such as the \cppinline{getQualifiedName} method were tried first. These small differences between node types and method names took a while to navigate through and because the information is needed both when creating the matchers and when doing the data extraction from the nodes this navigation is the most time-consuming part of creating a tool. It can also be quite hard to even identify which nodes are of interest, as there are so many different ones to choose from. In the end, the team developed the tools by figuring out what data was needed, then looking through the existing matchers/nodes in order to find a promising candidate and then trying the method out. This method is a systematic form of trial and error and caused some frustration in the process because it is so hard to identify the exact combination of node types and method calls to get the wanted information.